---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions, echo=FALSE, message = FALSE, warning = FALSE}
library("knitr") 
opts_chunk$set(echo=TRUE, message = FALSE, warning = FALSE, fig.height=6, fig.width=15, fig.path='figure/', cache.path='cache/')
```
## Loading and preprocessing the data

```{r results='hide'}
require("ggplot2")
require("plyr")

## unzip an load the data
unzip("activity.zip")
activity  <- read.csv("activity.csv", header = TRUE, sep = ",")
file.remove("activity.csv")
```

## What is mean total number of steps taken per day?
```{r }
## total number of steps taken per day
dailysteps <- ddply(activity, .(date), summarise, 
                total_steps = sum(steps, na.rm = TRUE))
```

```{r echo=FALSE}
mean_step = round(mean(dailysteps$total_steps))
median_step = median(dailysteps$total_steps)

## make a vector and a frame for the mean and the median line
ve <- c(mean_step, median_step)
df <- data.frame(ve)
```

```{r histogram, echo=FALSE}
## histogram of the total number of steps taken each day
histo <- ggplot(dailysteps, aes(x= total_steps)) +
          geom_histogram(fill="lightblue", colour="black", binwidth=2000) +
          geom_vline(data=df, aes(xintercept=ve), colour=factor(ve), linetype="dashed", size=1, show_guide = TRUE)+
          xlab("Total Number of steps") +
          ggtitle("The total number of steps taken each day")
histo
````

## Mean and median of the total number of steps taken per day
```{r cache=TRUE}
mean_step = round(mean(dailysteps$total_steps))
median_step = median(dailysteps$total_steps)
```
the **mean** (red line on histogram) is **`r format(mean_step, big.mark=" ")`** and the **median** (green line on histogram) is **`r format(median_step, big.mark=" ")`**. 

## What is the average daily activity pattern?
```{r}
## average number of steps by interval
average_steps <- ddply(activity, .(interval), summarise, 
                  mean_steps = mean(steps, na.rm = TRUE))
```

```{r echo=FALSE, cache=TRUE}
## the maximum number of steps
max_steps = max(average_steps$mean_steps)
cross_interval = average_steps[average_steps$mean_steps == max_steps,]$interval

```

```{r time_series_plot, echo=FALSE}
## Time series plot
series_plot <- ggplot( average_steps, aes(x=interval, y=mean_steps)) + 
                geom_line(colour = "red") +                                
                geom_vline(xintercept = cross_interval, colour="blue", linetype = "longdash") + 
                xlab("Time interval") +
                ylab("Average steps")+
                ggtitle("Average steps by time interval of a day")
series_plot
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r cache=TRUE}
## the maximum number of steps
max_steps = max(average_steps$mean_steps)
# the average cross interval
cross_interval = average_steps[average_steps$mean_steps == max_steps,]$interval
```
the average across interval (the blue line on the time series plot) is **`r cross_interval`** for **`r round(max_steps)`** maximum number of steps

## Imputing missing values
```{r}
## the total number of missing values in the dataset
total_na <- nrow(activity[is.na(activity$steps),])
```
there is **`r total_na`**  missing values in the data set.

### Filling the missing values in the data set

```{r}
## calcuate the mean steps for each interval
mean_interval_steps <- ddply(activity, .(interval), summarise, 
                    mean_steps = mean(steps, na.rm = TRUE))

## merge activity and interval mean steps
m <- join(x = activity, y = mean_interval_steps, by = "interval")
m[is.na(m$steps),]$steps = m[is.na(m$steps),]$mean_steps

activity2 <- m[, c(1:3)]
```

```{r histogramm2, echo=FALSE, cache=TRUE}
## total number of steps taken by day
dailysteps2 <- ddply(activity2, .(date), summarise, 
                    total_steps = sum(steps))

## Mean and median of the total number of steps taken per day
mean_step2 = round(mean(dailysteps2$total_steps))
median_step2 = round(median(dailysteps2$total_steps))

ve2 <- c(mean_step2, median_step2)
df2 <- data.frame(ve2)

## histogram of the total number of steps taken each day
histo2 <- ggplot(dailysteps2, aes(x= total_steps)) +
  geom_histogram(fill="lightblue", colour="black", binwidth=2000) +  
  geom_vline(aes(xintercept = mean_step2), color="red", linetype="dashed", size=1) +          
  geom_vline(aes(xintercept = median_step2), color="blue", linetype="dashed", size=1) +
  xlab("Total Number of steps") +
  ggtitle("The total number of steps taken each day")
histo2
```

### Mean and median
```{r cache=TRUE }
## Mean and median of the total number of steps taken per day
mean_step2 = round(mean(dailysteps2$total_steps))
median_step2 = round(median(dailysteps2$total_steps))
```
the **mean** ``r format(mean_step2, big.mark=" ")`` and the **median**  ``r format(median_step2, big.mark=" ")`` differs from the first estimated without empty values, they have the same value because of the filling of the NA values 

## Are there differences in activity patterns between weekdays and weekends?

### Create a new factor variable in the dataset with two levels "weekday" and "weekend"
```{r echo=FALSE, results='hide'}
## Set Locale English to avoid French days
Sys.setlocale("LC_TIME", "English")
```

```{r}
days<- weekdays(as.Date(activity$date, ))
activity$day = days

weekend_flag = days=="Sunday" | days=="Saturday"

activity$type[weekend_flag] = "weekend"
activity$type[!weekend_flag] = "weekday"


activity$day = as.factor(activity$type)
activity <- activity[, c(1:4)]
```


### Plot in a time series
```{r time_series_plot2, echo=FALSE}
## average number of steps by interval
new_average_steps <- ddply(activity, .(day, interval), summarise, 
                       mean_steps = mean(steps, na.rm = TRUE))

## Time series plot
series_plot2 <- ggplot( new_average_steps, aes(x=interval, y=mean_steps)) + 
                  geom_line(colour = "deepskyblue") +    
                  facet_wrap(~ day, ncol=1) +
                  theme_bw() +
                  theme(                  
                    panel.grid.major = element_blank(), 
                    panel.grid.minor = element_blank(), 
                    axis.line = element_line(colour = "black")) +                                     
                  xlab("Interval") +
                  ylab("Number of steps")

series_plot2
```